---
title: "Vis analysis"
author: "Group 8"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Setup and Libraries
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("rstatix", quietly = TRUE)) install.packages("rstatix")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")

library(dplyr)
library(tidyverse)
library(ggplot2)
library(readr)
library(rstatix)
library(readxl)
```


```{r data-loading and cleaning, include = FALSE}
# Load the dataset
sdg3 <- readxl::read_excel("xlsx/sdg_3.xlsx") 
sdg5 <- readxl::read_excel("xlsx/sdg_5.xlsx") 
sdg8 <- readxl::read_excel("xlsx/sdg_8.xlsx") 
sdg9 <- readxl::read_excel("xlsx/sdg_9.xlsx") 
sdg11 <- readxl::read_excel("xlsx/sdg_11.xlsx") 
sdg12 <- readxl::read_excel("xlsx/sdg_12.xlsx") 
sdg16 <- readxl::read_excel("xlsx/sdg_16.xlsx") 

# Inspect the data structure
str(sdg3)
str(sdg5)
str(sdg8)
str(sdg9)
str(sdg11)
str(sdg12)
str(sdg16)
summary(sdg3)
summary(sdg3)
summary(sdg5)
summary(sdg8)
summary(sdg9)
summary(sdg11)
summary(sdg12)
summary(sdg16)


# Convert
sdg3$year <- as.numeric(sdg3$year)
sdg3$bmi_value <- as.numeric(sdg3$bmi_value)
sdg3$bmi_value <- as.numeric(gsub("[^0-9.]", "", sdg3$bmi_value))  # Remove non-numeric characters




```


# SDG 3 - Good Health and Well-Being
## Ensure healthy lives and promote well-being for all at all ages
Sustainable Development Goal (SDG) 3 aims to ensure healthy lives and promote well-being for all at all ages. It focuses on reducing maternal and child mortality, combating infectious diseases (HIV, tuberculosis, malaria), promoting universal health coverage, improving mental health, and tackling non-communicable diseases (e.g., heart disease, diabetes).


We have decided to look specifically at BMI and how it has changed through the years.



```{r sdg 3, echo=FALSE}

filtered_sdg3 <- sdg3[sdg3$country %in% c("Sweden", "Norway", "Finland", "Denmark"), ]

ggplot(filtered_sdg3, aes(x = year, y = bmi_value, color = bmi, group = bmi)) +
  geom_line() + 
  geom_point() +
  facet_wrap(~ country) +  # Separate plots for each country
  labs(title = "BMI Category Trends Over the Years",
       x = "Year",
       y = "Percentage of Population",
       color = "BMI Category") +
  theme_minimal()

```

# SDG 5 - Gender Equality
## Achieve gender equality and empower all women and girls
Sustainable Development Goal (SDG) 5 aims to achieve gender equality and empower all women and girls. It focuses on eliminating discrimination, violence, and harmful practices (e.g., child marriage, female genital mutilation), ensuring equal opportunities in leadership and economic participation, and universal access to sexual and reproductive health rights.


We have decided to look at... 


# SDG 8 - Decent Work and Economic Growth
## Promote sustained, inclusive and sustainable economic growth, full and productive employment and decent work for all
Sustainable Development Goal (SDG) 8 aims to promote inclusive and sustainable economic growth, full and productive employment, and decent work for all. It focuses on:

* Sustained GDP growth and increased productivity
* Job creation and entrepreneurship
* Labor rights and safe working conditions
* Ending modern slavery, forced labor, and child labor
* Financial inclusion and equal access to economic resources


We have decided to look at... 


# SDG 9 - Industry, Innovation, and Infrastructure
## Build resilient infrastructure, promote inclusive and sustainable industrialization and foster innovation
Sustainable Development Goal (SDG) 9 aims to build resilient infrastructure, promote sustainable industrialization, and foster innovation. It focuses on:

* Expanding reliable and sustainable infrastructure (transport, energy, digital access).
* Supporting inclusive and sustainable industrialization with job creation.
* Increasing investment in R&D and innovation, especially in developing industries.
* Promoting clean technologies and environmentally friendly production.


We have decided to look at... 


# SDG 11 - Sustainable Cities and Communities
## Make cities and human settlements inclusive, safe, resilient and sustainable
Sustainable Development Goal (SDG) 11 aims to make cities inclusive, safe, resilient, and sustainable. It focuses on:

* Affordable housing and reducing slums.
* Sustainable urban transport and mobility.
* Reducing air pollution and environmental impact.
* Disaster risk reduction and climate resilience.
* Preserving cultural and natural heritage.


We have decided to look at... 


# SDG 12 - Responsible Consumption and Production
## Ensure sustainable consumption and production patterns
Sustainable Development Goal (SDG) 12 aims to ensure sustainable consumption and production patterns by:

* Reducing waste and pollution (food waste, plastic, industrial emissions).
* Promoting circular economy models (recycling, reusing, and reducing resource use).
* Encouraging sustainable business practices and corporate responsibility.
* Improving resource efficiency in production and supply chains.
* Raising awareness about sustainable consumption among individuals and businesses.


We have decided to look at... 



# SDG 16 - Peace, Justice, and Strong Institutions
## Promote peaceful and inclusive societies for sustainable development, provide access to justice for all and build effective, accountable and inclusive institutions at all levels
Sustainable Development Goal (SDG) 16 aims to promote peaceful and inclusive societies, provide access to justice for all, and build effective, accountable institutions. It focuses on:

* Reducing violence, corruption, and organized crime.
* Ensuring access to justice and human rights for all individuals.
* Strengthening democratic institutions and governance.
* Protecting fundamental freedoms and promoting transparency.
* Reducing illicit financial flows and bribery.


We have decided to look at... 



```{r session-info, include=FALSE}
# Session info for reproducibility
sessionInfo()
